Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-09-18 Thesis/Background.tex
More updates. On to final minutia

In this spirit, \citet{jeans_1915}\footnote{Notably before the infamous 1920 Curtis-Shapley debate \citep{curtis_shapley} establishing that so-called ``spiral nebulae" were actually external galaxies.} laid out a framework to describe galaxy evolution in terms of the \textit{moments} of the distribution function. Specifically, the idea is to multiply the CBE (Eq. \eqref{eq:cbe_cartesian}) by $\textbf{v}_i$ or $\textbf{v}_i \textbf{v}_j$ and integrate over the velocity part of phase space. These together yield the second-order Jeans equations \citep{BT},
\begin{eqnarray}
\deriv{\nu}{t} + \deriv{\nu \langle v_i \rangle}{x_i} &=& 0 \\
\nu \deriv{\langle v_j \rangle}{t} + \nu \langle v_i \rangle \deriv{\langle v_j \rangle}{x_i} &=& -\nu \deriv{\Phi}{x_j} - \deriv{\nu \sigma^2_{ij}}{x_i}\label{eq:jeans}
\end{eqnarray}
where $i,j \in (1,2,3)$, $\langle\cdot\rangle$ denotes expectation over $f$, $\mathbb{E}_f[\cdot]$, $\nu$ is the number density of particles, and $\sigma^2_{ij} = \langle v_i v_j \rangle - \langle v_i \rangle \langle v_j \rangle$ is the velocity dispersion tensor. In equilibrium for spherical systems, the second equation is often written as \citep{BT},
\begin{equation}
\deriv{\nu \langle v_r^2 \rangle}{r} + \nu \left(\deriv{\Phi}{r} + \frac{2 \langle v_r^2 \rangle - \langle v_\theta^2\rangle - \langle v_\phi^2 \rangle}{r} \right)
\end{equation}
where $\theta$ is the polar angle, $\phi$ is the azimuthal angle, and $r$ is the spherical radius\footnote{Note a slight abuse of notation. We said that $v_i$ has $i \in (1,2,3)$, but here we use symbols to represent specific coordinate systems. The point is that the indices have three unique values.}. Note that there is only one scalar equation. Symmetries in the velocity dispersion tensor for spherical systems require that the cross moments are zero \citep{BT}. For axisymmetric systems, the equilibrium Jeans equations are often written \citep{BT},
\begin{eqnarray}
\deriv{\nu \langle v_R^2 \rangle}{R} + \deriv{\nu \langle v_R v_z \rangle}{z} + \nu \left(\frac{\langle v_R^2 \rangle - \langle v_\phi^2 \rangle}{R} + \deriv{\Phi}{R} \right) &=& 0\\
\frac{1}{R} \deriv{R \nu \langle v_R v_z \rangle}{R} + \deriv{\nu \langle v_z^2 \rangle}{z}+ \nu \deriv{\Phi}{z} &=& 0\\
\frac{1}{R^2} \deriv{R^2 \nu \langle v_R v_\phi \rangle}{R} + \deriv{\nu \langle v_z v_\phi\rangle}{z} &=& 0
\end{eqnarray}
where $z$ is the Cartesian $z$, $R$ is the cylindrical radius, and $\phi$ is the polar angle. These equations look suspiciously like Euler's equations for fluid dynamics with a missing energy equation. \citet{jeans_1915} did not magically solve the statistical mechanical issues mentioned in \S\ref{sec:motivation}; given a number density and potential, we still have an unknown velocity dispersion tensor with six elements, and only 4 independent equations. In the simplified spherical case, we have three unknown elements and two independent equations.

>>>>

In this spirit, \citet{jeans_1915}\footnote{Notably before the famous 1920 Curtis-Shapley debate \citep{curtis_shapley} establishing that so-called ``spiral nebulae" were actually external galaxies.} laid out a framework to describe galaxy evolution in terms of the \textit{moments} of the distribution function. Specifically, the idea is to multiply the CBE (Eq. \eqref{eq:cbe_cartesian}) by $\textbf{v}_i$ or $\textbf{v}_i \textbf{v}_j$ and integrate over the velocity part of phase space. These together yield the second-order Jeans equations \citep{BT},
\begin{eqnarray}
\deriv{\nu}{t} + \deriv{\nu \langle v_i \rangle}{x_i} &=& 0 \\
\nu \deriv{\langle v_j \rangle}{t} + \nu \langle v_i \rangle \deriv{\langle v_j \rangle}{x_i} &=& -\nu \deriv{\Phi}{x_j} - \deriv{\nu \sigma^2_{ij}}{x_i}\label{eq:jeans}
\end{eqnarray}
where $i,j \in (1,2,3)$, $\langle\cdot\rangle$ denotes expectation over $f$, $\mathbb{E}_f[\cdot]$, $\nu$ is the number density of particles, and $\sigma^2_{ij} = \langle v_i v_j \rangle - \langle v_i \rangle \langle v_j \rangle$ is the velocity dispersion tensor. In equilibrium for spherical systems, the second equation is often written as \citep{BT},
\begin{equation}
\deriv{\nu \langle v_r^2 \rangle}{r} + \nu \left(\deriv{\Phi}{r} + \frac{2 \langle v_r^2 \rangle - \langle v_\theta^2\rangle - \langle v_\phi^2 \rangle}{r} \right)
\end{equation}
where $\theta$ is the polar angle, $\phi$ is the azimuthal angle, and $r$ is the spherical radius\footnote{Note a slight abuse of notation. We said that $v_i$ has $i \in (1,2,3)$, but here we use symbols to represent specific coordinate systems. The point is that the indices have three unique values.}. Note that there is only one scalar equation. Symmetries in the velocity dispersion tensor for spherical systems require that the cross moments are zero \citep{BT}. For axisymmetric systems, the equilibrium Jeans equations are often written \citep{BT},
\begin{eqnarray}
\deriv{\nu \langle v_R^2 \rangle}{R} + \deriv{\nu \langle v_R v_z \rangle}{z} + \nu \left(\frac{\langle v_R^2 \rangle - \langle v_\phi^2 \rangle}{R} + \deriv{\Phi}{R} \right) &=& 0\\
\frac{1}{R} \deriv{R \nu \langle v_R v_z \rangle}{R} + \deriv{\nu \langle v_z^2 \rangle}{z}+ \nu \deriv{\Phi}{z} &=& 0\\
\frac{1}{R^2} \deriv{R^2 \nu \langle v_R v_\phi \rangle}{R} + \deriv{\nu \langle v_z v_\phi\rangle}{z} &=& 0
\end{eqnarray}
where $z$ is the Cartesian $z$, $R$ is the cylindrical radius, and $\phi$ is the polar angle. These equations look suspiciously like Euler's equations for fluid dynamics with a missing energy equation. \citet{jeans_1915} did not magically solve the statistical mechanical issues mentioned in \S\ref{sec:motivation}; given a number density and potential, we still have an unknown velocity dispersion tensor with six elements, and only 4 independent equations. In the simplified spherical case, we have three unknown elements and two independent equations.
